Sparse spectral and $p$-finite element methods for partial differential equations on disk-slices and trapeziums

### 2019-05-23 papers/halfdisk/halfdisk.tex
Addressed SO comments - reformatted figs, changed labelling of P 1D OPs and added to proofs/lemmas

\newcommand{\half}{\frac{1}{2}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\C}{\mathbb{C}}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\No}{\mathbb{N}_0}
\newcommand{\Ylm}{Y^m_l}
\newcommand{\Ylmfull}{Y^m_l(\theta,\varphi)}
\newcommand{\Plm}{P^m_l}
\newcommand{\costheta}{\cos\theta}
\newcommand{\sintheta}{\sin\theta}
\newcommand{\cosphi}{\cos\varphi}
\newcommand{\sinphi}{\sin\varphi}
\newcommand{\eimphi}{e^{im\varphi}}
\newcommand{\alphalm}{\alpha^m_l}
\newcommand{\clm}{c^m_l}
\newcommand{\ctilde}{\tilde{c}^m_l}
\newcommand{\ctildemod}{\tilde{c}^{|m|}_l}
\newcommand{\chat}{\hat{c}^m_l}
\newcommand{\chatmod}{\hat{c}^{|m|}_l}
\newcommand{\ddx}{\frac{\mathrm{d}}{\mathrm{d}x}}
\newcommand{\ddy}{\frac{\mathrm{d}}{\mathrm{d}y}}
\newcommand{\pddx}{\frac{\partial}{\partial x}}
\newcommand{\pddy}{\frac{\partial}{\partial y}}
\newcommand{\pddn}{\frac{\partial}{\partial n}}
\newcommand{\dmdxm}{\frac{\mathrm{d}^m}{\mathrm{d}x^m}}

@@

\newcommand{\hdop}{H}
\newcommand{\bighdop}{\mathbb{\hdop}}
\newcommand{\hdopnk}{\hdop_{n,k}}
\newcommand{\hdopab}{\hdop^{(a,b)}}
\newcommand{\hdopnkab}{\hdop_{n,k}^{(a,b)}}
\newcommand{\Wab}{{W^{(a,b)}}}
\newcommand{\hdopmj}{\hdop_{m,j}}
\newcommand{\hdopmjab}{\hdop_{m,j}^{(a,b)}}
\newcommand{\alphaab}{\alpha^{(a,b)}}
\newcommand{\betaab}{\beta^{(a,b)}}
\newcommand{\bighdopab}{\bighdop^{(a,b)}}
\newcommand{\Dnt}{D^\top_n}
\newcommand{\Wii}{W^{(1,1)}}
\newcommand{\hdopii}{\hdop^{(1,1)}}
\newcommand{\bighdopii}{{\mathbb{\hdop}^{(1,1)}}}
\newcommand{\hdopoo}{\hdop^{(0,0)}}
\newcommand{\bighdopoo}{{\mathbb{\hdop}^{(0,0)}}}
\newcommand{\hdopoi}{\hdop^{(0,1)}}
\newcommand{\hdopio}{\hdop^{(1,0)}}
\newcommand{\genjac}{R}
\newcommand{\genjacnmk}{\genjac_{n-k}}
\newcommand{\genjacmmj}{\genjac_{m-j}}
\newcommand{\genjacw}{w_\genjac}
\newcommand{\normgenjac}{\omega_\genjac}
\newcommand{\normjac}{\omega_P}

@@

For disk slices and trapeziums, we  specialise  Proposition \ref{prop:construction} in the following definition.
\begin{definition}\label{def:construction}
Let
\begin{align*}
	\Omega := \{(x,y) \in \R^2 \quad | \quad \alpha < x < \beta, \: \gamma \rho(x) < y < \delta \rho(x)\}
\end{align*}
with $\rho : (\alpha, \beta) \to \R$. Further, let $\genjacw^{(a,b,c)}(x)$ and $w^{(a,b)}_P(x)$ be two weight functions on the intervals $(\alpha, \beta)$ and $(\gamma, \delta)$ respectively, given by:
\begin{align*}
\begin{cases}
\genjacw^{(a,b,c)}(x) &:= (\beta - x)^a \: (x - \alpha)^{b} \: \rho(x)^{c} \\
w^{(a,b)}_P(x) &:= (\delta-x)^a \: (x - \gamma)^b
\end{cases}
\end{align*}
and define the associated inner products by:
\begin{align}
	\ip< p, \: q >_{\genjacw^{(a,b,c)}} &:= \frac{1}{\normgenjac^{(a,b,c)}} \: \int_\alpha^\beta p(x) \: q(x) \: \genjacw^{(a,b,c)}(x) \: \D x \label{eqn:ipgenjac} \\
	\ip<p, q>_{w_P^{(a,b)}} &:= \frac{1}{\normjac^{(a,b)}} \: \int_\gamma^\delta p(y) \: q(y) \: w_P^{(a,b)}(y)\: \D y \label{eqn:ipjac}
\end{align}
where
\begin{align}
	\normgenjac^{(a,b,c)} := \int_\alpha^\beta \: \genjacw^{(a,b,c)}(x) \: \D x, \quad \normjac^{(a,b)} := \int_\gamma^\delta \: w_P^{(a,b)}(y) \: \D y. \label{eqn:ipnormalisations}
\end{align}
Further, denote the three-parameter family of orthonormal polynomials on $[\alpha,\beta]$ by $\{\genjac_n^{(a,b,c)}\}$, orthonormal with respect to the inner product defined in (\ref{eqn:ipgenjac}), and the two-parameter family of orthonormal polynomials on $[\gamma,\delta]$ by $\{P_n^{(a,b)}\}$, orthonormal with respect to the inner product defined in (\ref{eqn:ipjac}).

Assume that either of the following hold:
\begin{enumerate}
\item  \(\rho\) is a degree 1 polynomial, or 
\item \(\rho\) is the square root of a non-negative degree \(\le\) 2 polynomial, \(-\gamma = \delta > 0\), and \(w^{(a,b)}_P\) is an even function (i.e. $a = b$, and we can hence denote the weight as $w^{(a)}_P(x) = w^{(a,a)}_P(x) =  (\delta-x^2)^a$).
\end{enumerate}
We define the four-parameter 2D orthogonal polynomials on $\Omega$ by
\begin{align*}
	\hdopnk^{(a,b,c,d)}(x,y) := \genjacnmk^{(a, b, c+d+2k+1)}(x) \: \rho(x)^k \: P_k^{(d,c)}(\frac{y}{\rho(x)}), \quad (x,y) \in \Omega, 
\end{align*}
which are orthogonal with respect to the weight
\begin{align*}
	W^{(a,b,c,d)}(x,y) := \genjacw^{(a,b,c+d)}(x) \: w_P^{(d,c)}(\frac{y}{\rho(x)}), \quad (x,y) \in \Omega,
\end{align*}
that is,
\begin{align*}
	\ip< \hdopnk^{(a,b,c,d)}, \: \hdopmj^{(a,b,c,d)} >_{W^{(a,b,c,d)}} &= \normgenjac^{(a,b,c+d+2k+1)} \: \normjac^{(d,c)} \: \delta_{n,m} \: \delta_{k,j},
\end{align*}
where for $f, g : \Omega \to \R$ the inner product is defined as 
\begin{align*}
	\ip< f, \: g >_{W^{(a,b,c,d)}} := \iint_\Omega \: f(x,y) \: g(x,y) \: W^{(a,b,c,d)}(x,y) \: \D y \: \D x.
\end{align*}
\end{definition}

For the half disk, the weight $W^{(a,b)}(x,y) = x^a \: (1-x^2-y^2)^b$ results from setting:
\begin{align*}
\begin{cases}
(\alpha,\beta) &:= (0,1) \\
(\gamma,\delta) &:= (-1,1) \\
\rho(x) &:= (1-x^2)^{\half} \\
w_1(x) &:= \genjacw^{(a,b)}(x) := x^a \: (1-x^2)^b \\
w_2(x) &:= w_P^{(b)}(x) := (1-x^2)^b = (1-x)^b \: (1+x)^b,
\end{cases}
\end{align*}
Note here we have made the adjustment that $\genjacw^{(a,b,c)}(x) = (\beta - x)^a \: (x - \alpha)^{b} \: \rho(x)^{2c}$ and simply set the first parameter to zero and removed it for the specific half-disk case. This is for ease of presentiation, by ensuring we reduce the need to incorporate unnecessary parameters and instead yield a two-parameter family of 2D orthogonal polynomials.

The weight $w_P(x)$ is an ultraspherical weight, and the corresponding OPs are the normalized Jacobi polynomials  $\{\tilde P_n^{(b, b)}\}$ \sotodo{Change $P_n -> \tilde P_n$ everywhere}. The weight $w_R(x)$ is non-classical (it is in fact semi-classical, and is equivalent to a generalized Jacobi weight \cite[\S5]{magnus1995painleve}) and we introduce the notation $\{\genjac_n^{(a, b)}\}$ for the orthogonal polynomials. The two 1D families of orthogonal polynomials we have denoted $\{\genjac_{n}^{(a,b)}\}$ and $\{P_n^{(b,b)}\}$ are orthonormal, with respect to the inner products respectively given by:
\begin{align*}
	\ip< p, \: q >_{\genjacw^{(a,b)}} &:= \frac{1}{\normgenjac^{(a,b)}} \: \int_0^1 p(x) \: q(x) \: \genjacw^{(a,b)}(x) \: \D x, \\
	\ip<p, q>_{w_P^{(b)}} &:= \frac{1}{\normjac^{(b)}} \: \int_{-1}^1 p(y) \: q(y) \: w_P^{(b)}(y)\: \D y,
\end{align*}
i.e. $\norm{\genjac_{n}^{(a,b)}}_{\genjacw^{(a,b)}} = 1$ $\forall n = 0,1,2,\dots$ and $\norm{P_{n}^{(b,b)}}_{w_P^{(b)}} = 1$ $\forall n = 0,1,2,\dots$. For clarity, we use the notation that $\norm{p}_{\genjacw^{(a,b)}}^2 := \ip<p, p>_{\genjacw^{(a,b)}}$ and that $\norm{p}_{w_P^{(b)}}^2 := \ip<p, p>_{w_P^{(b)}}$.

Thus we arrive at a two parameter family of 2D orthogonal polynomials $\{\hdopnkab\}$ given by, for $0 \le k \le n, \: n = 0,1,2,\dots,$
\begin{align}\label{eq:diskpolys}
	\hdopnk^{(a,b)}(x,y) := \genjacnmk^{(a, b+k+\half)}(x) \: \rho(x)^k \: P_k^{(b,b)}\fpr(\frac{y}{\rho(x)}), \quad (x,y) \in \Omega, 
\end{align}
orthogonal on $\Omega$ with respect to the weight
\begin{align*}
	W^{(a,b)}(x,y) = \genjacw^{(a,b)}(x) \: w_P^{(b)}\fpr(\frac{y}{\rho(x)}) =  \frac{x^a \: (1-x^2-y^2)^b}{\normgenjac^{(a,b)} \normjac^{(b)}}.
\end{align*}
That is, they are orthogonal (for given parameters $a,b$) with respect to the inner product given by
\begin{align*}
	\ip< \hdopnkab, \: \hdopmjab >_\Wab &=  \iint_\Omega \: \hdopnkab(x,y) \: \hdopmjab(x,y) \: \Wab(x,y) \: \D y \: \D x.
\end{align*}
We can see that they are indeed orthogonal using the change of variable $t = \frac{y}{\rho(x)}$:
\begin{align}
	\ip< \hdopnkab, \: \hdopmjab >_\Wab &=  \iint_\Omega \: \Big[ \genjacnmk^{(a,b+k+\half)}(x) \: \genjacmmj^{(a,b+j+\half)}(x) \: \rho(x)^{k+j}  \nonumber \\
		&\quad \quad \quad \quad \quad \quad \cdot \: P_k^{(b,b)}\fpr(\frac{y}{\rho(x)}) \: P_j^{(b,b)}\fpr(\frac{y}{\rho(x)})  \: \Wab(x,y) \Big] \: \D y \: \D x \nonumber \\
	&= \Big( \int_\alpha^\beta \: {\genjacnmk^{(a,b+k+\half)}}(x) \: {\genjacmmj^{(a,b+j+\half)}}(x) \: \genjacw^{(a,b+\half(k+j+1))}(x) \: \D x \Big) \nonumber \\
	&\quad \quad \quad \quad \quad \quad \cdot \: \Big( \int_\gamma^\delta \: {P_k^{(b,b)}}(t) \:  {P_j^{(b,b)}}(t) \: w_P^{(b)}(t) \: \D t \Big) \nonumber \\
	&= \normjac^{(b)} \: \delta_{k,j} \: \int_\alpha^\beta \: {\genjacnmk^{(a,b+k+\half)}}(x) \: {\genjac_{m-k}^{(a,b+k+\half)}}(x) \: \genjacw^{(a,b+k+\half)}(x) \: \D x \nonumber \\ 
	&= \normgenjac^{(a,b+k+\half)} \: \normjac^{(b)} \: \delta_{n,m} \: \delta_{k,j}. \label{eqn:normhdop}
\end{align}
Again, for clarity we will use the notation that for $p : \Omega \to \R$, $\norm{p}_{\Wab}^2 := \ip<p, p>_{\Wab}$.

@@

We can express the three-term recurrences associated to $\genjac_n^{(a,b)}$ and $P_n^{(b,b)}$ as 
\begin{align}
	x \genjac_n^{(a,b)}(x) &= \beta_n^{(a,b)} \genjac_{n+1}^{(a,b)}(x) + \alpha_n^{(a,b)} \genjac_n^{(a,b)}(x) + \beta_{n-1}^{(a,b)} \genjac_{n-1}^{(a,b)}(x) \label{eqn:Hrecurrence} \\
	y P_n^{(b,b)}(y) &= \delta_n^{(b)} P_{n+1}^{(b,b)}(y) + \gamma_n^{(b)} P_n^{(b,b)}(y) + \delta_{n-1}^{(b)} P_{n-1}^{(b,b)}(y),
\end{align}
where we note that \(\gamma_n^{(b)} = 0\), for \( n = 0,1,2,\dots\). We can use these to determine the 2D recurrences for $\hdopnkab(x,y)$. Importantly, we can deduce sparsity in the recurrence relationships:

@@

\begin{proof}
The 3-term recurrence for multiplication by $x$ follows from equation (\ref{eqn:Hrecurrence}). For the recurrence for multiplication by $y$, since $\{\hdopmjab\}$ for $m = 0,\dots,n+1$, $j = 0,\dots,m$ is an orthogonal basis for any degree $n+1$ polynomial, we can expand $y \: \hdopnkab(x,) = \sum_{m=0}^{n+1} \sum_{j=0}^m c_{m,j} \: \hdopmjab(x,y)$. These coefficients are given by
\begin{align*}
	c_{m,j} = {\ip< y \: \hdopnkab, \hdopmjab >_{\Wab}}{\norm{\hdopmjab}^{-2}_{\Wab}}.
\end{align*}
Recall from equation (\ref{eqn:normhdop}) that $\norm{\hdopmjab}_{\Wab}^2 = \normgenjac^{(a,b+j+\half)} \: \normjac^{(b)}$. Then for $m = 0,\dots,n+1$, $j = 0,\dots,m$, using the change of variable $t = \frac{y}{\rho(x)}$:
\begin{align*}
	\ip<y \hdopnkab, \hdopmjab>_\Wab &=  \iint_\Omega \hdopnkab(x,y) \: \hdopmjab(x,y) \: y \: \Wab(x,y) \: dy \: dx \\
	&= \Big( \int^1_0 \genjacnmk^{(a, b+k+\half)}(x) \: \genjacmmj^{(a, b+j+\half)}(x) \: \rho(x)^{k+j+2} \: \genjacw^{(a, b)}(x) \: \D x \Big) \nonumber \\
	& \quad \quad \quad\cdot \: \Big( \int^1_{-1} P_k^{(b,b)}(t) \: P_j^{(b,b)}(t) \: t \: w_P^{(b)}(t) \: \D t \Big) \\
	&= \Big( \int^1_0 \genjacnmk^{(a, b+k+\half)}(x) \: \genjacmmj^{(a, b+j+\half)}(x) \: \genjacw^{(a, b+1+\half(k+j))}(x) \: \D x \Big) \nonumber \\
	& \quad \quad \quad\cdot \: \Big( \int^1_{-1} P_k^{(b,b)}(t) \: P_j^{(b,b)}(t) \: t \: w_P^{(b)}(t) \: \D t \Big)
\end{align*}
\begin{align*}
	=
	\begin{cases}
    		\delta_k^{(b)} \: \normjac^{(b)} \: \normgenjac^{(a,b+k+\frac{3}{2})} \: \ip<\genjacnmk^{(a, b+k+\half)}, \genjac_{m-k-1}^{(a, b+k+\frac{3}{2})}>_{\genjacw^{(a, b+k+\frac{3}{2})}} \quad& \text{if } j = k+1 \\
		\delta_{k-1}^{(b)} \: \normjac^{(b)} \: \normgenjac^{(a,b+k-\half)} \: \ip<\genjacnmk^{(a, b+k+\half)}, \: \rho(x)^2 \genjac_{m-k+1}^{(a, b+k-\half)}>_{\genjacw^{(a, b+k-\half)}} \quad& \text{if } j = k-1 \\
		0 & \text{otherwise}
      	\end{cases}
\end{align*}
where $\ip<\genjacnmk^{(a, b+k+\half)}, \genjac_{m-k-1}^{(a, b+k+\frac{3}{2})}>_{\genjacw^{(a, b+k+\frac{3}{2})}}$ and $\ip<\genjacnmk^{(a, b+k+\half)}, \: \rho(x)^2 \genjac_{m-k+1}^{(a, b+k-\half)}>_{\genjacw^{(a, b+k-\half)}}$ are zero for $m < n-1$.

>>>>

\newcommand{\bigO}{\mathcal{O}}
\newcommand{\half}{\frac{1}{2}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\C}{\mathbb{C}}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\No}{\mathbb{N}_0}
\newcommand{\Ylm}{Y^m_l}
\newcommand{\Ylmfull}{Y^m_l(\theta,\varphi)}
\newcommand{\Plm}{\jac^m_l}
\newcommand{\costheta}{\cos\theta}
\newcommand{\sintheta}{\sin\theta}
\newcommand{\cosphi}{\cos\varphi}
\newcommand{\sinphi}{\sin\varphi}
\newcommand{\eimphi}{e^{im\varphi}}
\newcommand{\alphalm}{\alpha^m_l}
\newcommand{\clm}{c^m_l}
\newcommand{\ctilde}{\tilde{c}^m_l}
\newcommand{\ctildemod}{\tilde{c}^{|m|}_l}
\newcommand{\chat}{\hat{c}^m_l}
\newcommand{\chatmod}{\hat{c}^{|m|}_l}
\newcommand{\ddx}{\frac{\mathrm{d}}{\mathrm{d}x}}
\newcommand{\ddy}{\frac{\mathrm{d}}{\mathrm{d}y}}
\newcommand{\pddx}{\frac{\partial}{\partial x}}
\newcommand{\pddy}{\frac{\partial}{\partial y}}
\newcommand{\pddn}{\frac{\partial}{\partial n}}
\newcommand{\dmdxm}{\frac{\mathrm{d}^m}{\mathrm{d}x^m}}

@@

\newcommand{\hdop}{H}
\newcommand{\bighdop}{\mathbb{\hdop}}
\newcommand{\hdopnk}{\hdop_{n,k}}
\newcommand{\hdopab}{\hdop^{(a,b)}}
\newcommand{\hdopnkab}{\hdop_{n,k}^{(a,b)}}
\newcommand{\Wab}{{W^{(a,b)}}}
\newcommand{\hdopmj}{\hdop_{m,j}}
\newcommand{\hdopmjab}{\hdop_{m,j}^{(a,b)}}
\newcommand{\alphaab}{\alpha^{(a,b)}}
\newcommand{\betaab}{\beta^{(a,b)}}
\newcommand{\bighdopab}{\bighdop^{(a,b)}}
\newcommand{\Dnt}{D^\top_n}
\newcommand{\Wii}{W^{(1,1)}}
\newcommand{\hdopii}{\hdop^{(1,1)}}
\newcommand{\bighdopii}{{\mathbb{\hdop}^{(1,1)}}}
\newcommand{\hdopoo}{\hdop^{(0,0)}}
\newcommand{\bighdopoo}{{\mathbb{\hdop}^{(0,0)}}}
\newcommand{\hdopoi}{\hdop^{(0,1)}}
\newcommand{\hdopio}{\hdop^{(1,0)}}
\newcommand{\jac}{{\tilde P}}
\newcommand{\genjac}{R}
\newcommand{\genjacnmk}{\genjac_{n-k}}
\newcommand{\genjacmmj}{\genjac_{m-j}}
\newcommand{\genjacw}{w_\genjac}
\newcommand{\jacw}{w_\jac}
\newcommand{\normgenjac}{\omega_\genjac}
\newcommand{\normjac}{\omega_\jac}

@@

For disk slices and trapeziums, we specialise Proposition \ref{prop:construction} in the following definition.
\begin{definition}\label{def:construction}
Let
\begin{align*}
	\Omega := \{(x,y) \in \R^2 \quad | \quad \alpha < x < \beta, \: \gamma \rho(x) < y < \delta \rho(x)\}
\end{align*}
with $\rho : (\alpha, \beta) \to \R$. Further, let $\genjacw^{(a,b,c)}(x)$ and $\jacw^{(a,b)}(x)$ be two weight functions on the intervals $(\alpha, \beta)$ and $(\gamma, \delta)$ respectively, given by:
\begin{align*}
\begin{cases}
\genjacw^{(a,b,c)}(x) &:= (\beta - x)^a \: (x - \alpha)^{b} \: \rho(x)^{c} \\
\jacw^{(a,b)}(x) &:= (\delta-x)^a \: (x - \gamma)^b
\end{cases}
\end{align*}
and define the associated inner products by:
\begin{align}
	\ip< p, \: q >_{\genjacw^{(a,b,c)}} &:= \frac{1}{\normgenjac^{(a,b,c)}} \: \int_\alpha^\beta p(x) \: q(x) \: \genjacw^{(a,b,c)}(x) \: \D x \label{eqn:ipgenjac} \\
	\ip<p, q>_{\jacw^{(a,b)}} &:= \frac{1}{\normjac^{(a,b)}} \: \int_\gamma^\delta p(y) \: q(y) \: \jacw^{(a,b)}(y)\: \D y \label{eqn:ipjac}
\end{align}
where
\begin{align}
	\normgenjac^{(a,b,c)} := \int_\alpha^\beta \: \genjacw^{(a,b,c)}(x) \: \D x, \quad \normjac^{(a,b)} := \int_\gamma^\delta \: \jacw^{(a,b)}(y) \: \D y. \label{eqn:ipnormalisations}
\end{align}
Further, denote the three-parameter family of orthonormal polynomials on $[\alpha,\beta]$ by $\{\genjac_n^{(a,b,c)}\}$, orthonormal with respect to the inner product defined in (\ref{eqn:ipgenjac}), and the two-parameter family of orthonormal polynomials on $[\gamma,\delta]$ by $\{\jac_n^{(a,b)}\}$, orthonormal with respect to the inner product defined in (\ref{eqn:ipjac}).

Assume that either of the following hold:
\begin{enumerate}
\item  \(\rho\) is a degree 1 polynomial, or 
\item \(\rho\) is the square root of a non-negative degree \(\le\) 2 polynomial, \(-\gamma = \delta > 0\), and \(\jacw^{(a,b)}\) is an even function (i.e. $a = b$, and we can hence denote the weight as $\jacw^{(a)}(x) = w^{(a,a)}_P(x) =  (\delta-x^2)^a$).
\end{enumerate}
We define the four-parameter 2D orthogonal polynomials on $\Omega$ by
\begin{align*}
	\hdopnk^{(a,b,c,d)}(x,y) := \genjacnmk^{(a, b, c+d+2k+1)}(x) \: \rho(x)^k \: \jac_k^{(d,c)}(\frac{y}{\rho(x)}), \quad (x,y) \in \Omega, 
\end{align*}
which are orthogonal with respect to the weight
\begin{align*}
	W^{(a,b,c,d)}(x,y) := \genjacw^{(a,b,c+d)}(x) \: \jacw^{(d,c)}(\frac{y}{\rho(x)}), \quad (x,y) \in \Omega,
\end{align*}
that is,
\begin{align*}
	\ip< \hdopnk^{(a,b,c,d)}, \: \hdopmj^{(a,b,c,d)} >_{W^{(a,b,c,d)}} &= \normgenjac^{(a,b,c+d+2k+1)} \: \normjac^{(d,c)} \: \delta_{n,m} \: \delta_{k,j},
\end{align*}
where for $f, g : \Omega \to \R$ the inner product is defined as 
\begin{align*}
	\ip< f, \: g >_{W^{(a,b,c,d)}} := \iint_\Omega \: f(x,y) \: g(x,y) \: W^{(a,b,c,d)}(x,y) \: \D y \: \D x.
\end{align*}
\end{definition}

For the half disk, the weight $W^{(a,b)}(x,y) = x^a \: (1-x^2-y^2)^b$ results from setting:
\begin{align*}
\begin{cases}
(\alpha,\beta) &:= (0,1) \\
(\gamma,\delta) &:= (-1,1) \\
\rho(x) &:= (1-x^2)^{\half} \\
w_1(x) &:= \genjacw^{(a,b)}(x) := x^a \: (1-x^2)^b \\
w_2(x) &:= \jacw^{(b)}(x) := (1-x^2)^b = (1-x)^b \: (1+x)^b,
\end{cases}
\end{align*}
Note here we have made the adjustment that $\genjacw^{(a,b,c)}(x) = (\beta - x)^a \: (x - \alpha)^{b} \: \rho(x)^{2c}$ and simply set the first parameter to zero and removed it for the specific half-disk case. This is for ease of presentiation, by ensuring we reduce the need to incorporate unnecessary parameters and instead yield a two-parameter family of 2D orthogonal polynomials.

The weight $\jacw(x)$ is an ultraspherical weight, and the corresponding OPs are the normalized Jacobi polynomials $\jac_n^{(b, b)}\}$. The weight $w_R(x)$ is non-classical (it is in fact semi-classical, and is equivalent to a generalized Jacobi weight \cite[\S5]{magnus1995painleve}) and we introduce the notation $\{\genjac_n^{(a, b)}\}$ for the orthogonal polynomials. The two 1D families of orthogonal polynomials we have denoted $\{\genjac_{n}^{(a,b)}\}$ and $\{\jac_n^{(b,b)}\}$ are orthonormal, with respect to the inner products respectively given by:
\begin{align*}
	\ip< p, \: q >_{\genjacw^{(a,b)}} &:= \frac{1}{\normgenjac^{(a,b)}} \: \int_0^1 p(x) \: q(x) \: \genjacw^{(a,b)}(x) \: \D x, \\
	\ip<p, q>_{\jacw^{(b)}} &:= \frac{1}{\normjac^{(b)}} \: \int_{-1}^1 p(y) \: q(y) \: \jacw^{(b)}(y)\: \D y,
\end{align*}
i.e. $\norm{\genjac_{n}^{(a,b)}}_{\genjacw^{(a,b)}} = 1$ $\forall n = 0,1,2,\dots$ and $\norm{\jac_{n}^{(b,b)}}_{\jacw^{(b)}} = 1$ $\forall n = 0,1,2,\dots$. For clarity, we use the notation that $\norm{p}_{\genjacw^{(a,b)}}^2 := \ip<p, p>_{\genjacw^{(a,b)}}$ and that $\norm{p}_{\jacw^{(b)}}^2 := \ip<p, p>_{\jacw^{(b)}}$.

Thus we arrive at a two parameter family of 2D orthogonal polynomials $\{\hdopnkab\}$ given by, for $0 \le k \le n, \: n = 0,1,2,\dots,$
\begin{align}\label{eq:diskpolys}
	\hdopnk^{(a,b)}(x,y) := \genjacnmk^{(a, b+k+\half)}(x) \: \rho(x)^k \: \jac_k^{(b,b)}\fpr(\frac{y}{\rho(x)}), \quad (x,y) \in \Omega, 
\end{align}
orthogonal on $\Omega$ with respect to the weight
\begin{align*}
	W^{(a,b)}(x,y) = \genjacw^{(a,b)}(x) \: \jacw^{(b)}\fpr(\frac{y}{\rho(x)}) =  \frac{x^a \: (1-x^2-y^2)^b}{\normgenjac^{(a,b)} \normjac^{(b)}}.
\end{align*}
That is, they are orthogonal (for given parameters $a,b$) with respect to the inner product given by
\begin{align*}
	\ip< \hdopnkab, \: \hdopmjab >_\Wab &=  \iint_\Omega \: \hdopnkab(x,y) \: \hdopmjab(x,y) \: \Wab(x,y) \: \D y \: \D x.
\end{align*}
We can see that they are indeed orthogonal using the change of variable $t = \frac{y}{\rho(x)}$:
\begin{align}
	\ip< \hdopnkab, \: \hdopmjab >_\Wab &=  \iint_\Omega \: \Big[ \genjacnmk^{(a,b+k+\half)}(x) \: \genjacmmj^{(a,b+j+\half)}(x) \: \rho(x)^{k+j}  \nonumber \\
		&\quad \quad \quad \quad \quad \quad \cdot \: \jac_k^{(b,b)}\fpr(\frac{y}{\rho(x)}) \: \jac_j^{(b,b)}\fpr(\frac{y}{\rho(x)})  \: \Wab(x,y) \Big] \: \D y \: \D x \nonumber \\
	&= \Big( \int_\alpha^\beta \: {\genjacnmk^{(a,b+k+\half)}}(x) \: {\genjacmmj^{(a,b+j+\half)}}(x) \: \genjacw^{(a,b+\half(k+j+1))}(x) \: \D x \Big) \nonumber \\
	&\quad \quad \quad \quad \quad \quad \cdot \: \Big( \int_\gamma^\delta \: {\jac_k^{(b,b)}}(t) \:  {\jac_j^{(b,b)}}(t) \: \jacw^{(b)}(t) \: \D t \Big) \nonumber \\
	&= \normjac^{(b)} \: \delta_{k,j} \: \int_\alpha^\beta \: {\genjacnmk^{(a,b+k+\half)}}(x) \: {\genjac_{m-k}^{(a,b+k+\half)}}(x) \: \genjacw^{(a,b+k+\half)}(x) \: \D x \nonumber \\ 
	&= \normgenjac^{(a,b+k+\half)} \: \normjac^{(b)} \: \delta_{n,m} \: \delta_{k,j}. \label{eqn:normhdop}
\end{align}
Again, for clarity we will use the notation that for $p : \Omega \to \R$, $\norm{p}_{\Wab}^2 := \ip<p, p>_{\Wab}$.

@@

We can express the three-term recurrences associated to $\genjac_n^{(a,b)}$ and $\jac_n^{(b,b)}$ as 
\begin{align}
	x \genjac_n^{(a,b)}(x) &= \beta_n^{(a,b)} \genjac_{n+1}^{(a,b)}(x) + \alpha_n^{(a,b)} \genjac_n^{(a,b)}(x) + \beta_{n-1}^{(a,b)} \genjac_{n-1}^{(a,b)}(x) \label{eqn:Hrecurrence} \\
	y \jac_n^{(b,b)}(y) &= \delta_n^{(b)} \jac_{n+1}^{(b,b)}(y) + \gamma_n^{(b)} \jac_n^{(b,b)}(y) + \delta_{n-1}^{(b)} \jac_{n-1}^{(b,b)}(y),
\end{align}
where we note that \(\gamma_n^{(b)} = 0\), for \( n = 0,1,2,\dots\). We can use these to determine the 2D recurrences for $\hdopnkab(x,y)$. Importantly, we can deduce sparsity in the recurrence relationships:

@@

\begin{proof}
The 3-term recurrence for multiplication by $x$ follows from equation (\ref{eqn:Hrecurrence}). For the recurrence for multiplication by $y$, since $\{\hdopmjab\}$ for $m = 0,\dots,n+1$, $j = 0,\dots,m$ is an orthogonal basis for any degree $n+1$ polynomial, we can expand $y \: \hdopnkab(x,) = \sum_{m=0}^{n+1} \sum_{j=0}^m c_{m,j} \: \hdopmjab(x,y)$. These coefficients are given by
\begin{align*}
	c_{m,j} = {\ip< y \: \hdopnkab, \hdopmjab >_{\Wab}}{\norm{\hdopmjab}^{-2}_{\Wab}}.
\end{align*}
Recall from equation (\ref{eqn:normhdop}) that $\norm{\hdopmjab}_{\Wab}^2 = \normgenjac^{(a,b+j+\half)} \: \normjac^{(b)}$. Then for $m = 0,\dots,n+1$, $j = 0,\dots,m$, using the change of variable $t = \frac{y}{\rho(x)}$:
\begin{align*}
	\ip<y \hdopnkab, \hdopmjab>_\Wab &=  \iint_\Omega \hdopnkab(x,y) \: \hdopmjab(x,y) \: y \: \Wab(x,y) \: dy \: dx \\
	&= \Big( \int^1_0 \genjacnmk^{(a, b+k+\half)}(x) \: \genjacmmj^{(a, b+j+\half)}(x) \: \rho(x)^{k+j+2} \: \genjacw^{(a, b)}(x) \: \D x \Big) \nonumber \\
	& \quad \quad \quad\cdot \: \Big( \int^1_{-1} \jac_k^{(b,b)}(t) \: \jac_j^{(b,b)}(t) \: t \: \jacw^{(b)}(t) \: \D t \Big) \\
	&= \Big( \int^1_0 \genjacnmk^{(a, b+k+\half)}(x) \: \genjacmmj^{(a, b+j+\half)}(x) \: \genjacw^{(a, b+1+\half(k+j))}(x) \: \D x \Big) \nonumber \\
	& \quad \quad \quad\cdot \: \Big( \int^1_{-1} \jac_k^{(b,b)}(t) \: \jac_j^{(b,b)}(t) \: t \: \jacw^{(b)}(t) \: \D t \Big)
\end{align*}
\begin{align*}
	=
	\begin{cases}
    		\delta_k^{(b)} \: \normjac^{(b)} \: \normgenjac^{(a,b+k+\frac{3}{2})} \: \ip<\genjacnmk^{(a, b+k+\half)}, \genjac_{m-k-1}^{(a, b+k+\frac{3}{2})}>_{\genjacw^{(a, b+k+\frac{3}{2})}} \quad& \text{if } j = k+1 \\
		\delta_{k-1}^{(b)} \: \normjac^{(b)} \: \normgenjac^{(a,b+k-\half)} \: \ip<\genjacnmk^{(a, b+k+\half)}, \: \rho(x)^2 \genjac_{m-k+1}^{(a, b+k-\half)}>_{\genjacw^{(a, b+k-\half)}} \quad& \text{if } j = k-1 \\
		0 & \text{otherwise}
      	\end{cases}
\end{align*}
where $\ip<\genjacnmk^{(a, b+k+\half)}, \genjac_{m-k-1}^{(a, b+k+\frac{3}{2})}>_{\genjacw^{(a, b+k+\frac{3}{2})}}$ and $\ip<\genjacnmk^{(a, b+k+\half)}, \: \rho(x)^2 \genjac_{m-k+1}^{(a, b+k-\half)}>_{\genjacw^{(a, b+k-\half)}}$ are zero for $m < n-1$.
